---
title: "Week 1 Exercises"
author: "Natasha Becker Bertelsen" 
date: 2/2-2022
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Setting working directory
setwd('.')

# Loading packages
pacman::p_load(tidyverse, ggplot2, ggfortify, dplyr, stringr, cowplot, lmerTest, DescTools)

```

1.2 Sketching a regression model and data: Figure 1.1b shows data corresponding to the fitted line y = 46.3 + 3.0x with residual standard deviation 3.9, and values of x ranging roughly from 0 to 4%. 

a) Sketch hypothetical data with the same range of x but corresponding to the line y = 30 + 10x with residual standard deviation 3.9.

```{r}
# Simulate data points for x 
x <- rnorm(1000, mean = 2, sd = 0.6)
x
hist(x)

# Simulate residuals
error <- rnorm(1000, mean = 0, sd = 3.9)
error
hist(error)

# Fit new data into model to find the corresponding y-values
y = 30 + 10 * x + error

# Create data frame from x and y values
df <- data.frame(x, y)

# Fitting regression model
model1 <- lm(y ~ x, data = df)
summary(model1)
autoplot(lm(y ~ x, data = df))

```

b) Sketch hypothetical data with the same range of x but corresponding to the line y = 30 + 10x with residual standard deviation 10. 

```{r}
# Simulate data points for x 
x1 <- rnorm(900, mean = 2, sd = 0.6)
x1
hist(x1)

# Simulate residuals
error1 <- rnorm(900, mean = 0, sd = 10)
error1
hist(error1)

# Fit new data into model to find the corresponding y-values
y1 = 30 + 10 * x1 + error1

# Create data frame with x and y values
df1 <- data.frame(x1, y1)

# Fitting model
model2 <- lm(y1 ~ x1, data = df1)
summary(model2)
autoplot(lm(y1 ~ x1, data = df1))
```

2. Applying learned skills on your selected data set:

2.1. Use a descriptive statistics and visual investigation to give a preliminary look into the data. (Use some of the things from chapter 2)
```{r}
# Load in data set 
df_epilepsy <- read.csv("epilepsy.csv")
```

Descriptive statistics investigation
```{r}
# Inspect first 10 rows of the data frame 
head(df_epilepsy,10)

# Check the numbers of rows and columns
nrow(df_epilepsy)
ncol(df_epilepsy)

# Check column names
colnames(df_epilepsy)

# Check summary statistics 
summary(df_epilepsy)

# Check structure of data frame 
str(df_epilepsy)

```

``` 
> Treatment: factor with levels placebo progabide indicating whether the anti-epilepsy drug Progabide has been applied or not, column-type CATEGORICAL COVARIATE.
> Base: number of epileptic attacks recorded during 8 week period prior to randomization, column-type CONTINUOUS COVARIATE.
> Age: age of the patients, column-type CONTINUOUS COVARIATE.
> Seizure_rate: number of epilepsy attacks patients have during the follow-up period, column-type OBSERVATION.
> Period: measurement period, column-type TIME.
> Subject: patient identification number, column-type ID.
```

Visual investigation 
```{r}
# Scatter plot seizure rate, period and treatment 
df_epilepsy %>% 
  ggplot() +
  aes(x = period, y = seizure.rate, color = treatment) +
  geom_point() +
  geom_jitter(width = .4, height = .4) +
  ggtitle("Scatterplot") +
  labs(x='Period',y ='Seizure rate')

# Histogram seizure rate
df_epilepsy %>% 
  ggplot() +
  aes(x=seizure.rate, fill = "peachpuff") +
  geom_histogram(binwidth = 0.5) +
  theme_minimal() +
  ggtitle("Histogram of seizure rate") +
  theme_minimal()

# Histogram log-transformed seizure rate
df_epilepsy %>% 
  ggplot() +
  aes(x=log(seizure.rate), fill = "peachpuff") +
  geom_histogram(binwidth = 0.5) +
  theme_minimal() +
  ggtitle("Histogram of seizure rate") +
  theme_minimal()

# QQ-plot seizure rate
qqnorm(df_epilepsy$seizure.rate, main = "QQ-plot Seizure rate")
qqline(df_epilepsy$seizure.rate)

# QQ-plot log-transformed seizure rate - why does it not work? 
qqnorm(log(df_epilepsy$seizure.rate), main = "Log transformed seizure rate\n Normal Q-Q Plot")
qqline(log(df_epilepsy$seizure.rate))

# Histogram base rate
df_epilepsy %>% 
  ggplot() +
  aes(x=base, fill = "peachpuff") +
  geom_histogram(binwidth = 5) +
  theme_minimal() +
  ggtitle("Histogram of base") +
  theme_minimal()

# Log-transformed histogram base rate
df_epilepsy %>% 
  ggplot() +
  aes(x=log(base), fill = "peachpuff") +
  geom_histogram(binwidth = 0.5) +
  theme_minimal() +
  ggtitle("Histogram of log-transformed base") +
  theme_minimal()

# QQ-plot 
qqnorm(df_epilepsy$base, main = "QQ-plot Seizure rate")
qqline(df_epilepsy$base)

# QQ-plot log-transformed base
qqnorm(log(df_epilepsy$base), main = "Log transformed seizure rate\n Normal Q-Q Plot")
qqline(log(df_epilepsy$base))

# Histogram of age 
df_epilepsy %>% 
  ggplot() +
  aes(x=age, fill = "peachpuff") +
  geom_histogram(binwidth = 1) +
  theme_minimal() +
  ggtitle("Histogram of age") +
  theme_minimal()

```

2.2. Create several models and discuss what each model reliably can infer anything about. Mixed effect, which random effect? Logistic regression? Interaction effects? Should we dummy code some variables?
```{r}
# Making treatment a factor variable (1 and 2 instead of placebo and progabide
df_epilepsy$treatment<-as.factor(df_epilepsy$treatment)
df_epilepsy$treatment<-as.numeric(df_epilepsy$treatment)
df_epilepsy$treatment<-as.factor(df_epilepsy$treatment)
levels(df_epilepsy$treatment)

# Logistic model
logistic_model <- glm(treatment ~ seizure.rate, family = binomial, data = df_epilepsy)
summary(logistic_model)

# Mixed-effects models 
intercept_model <- lmerTest::lmer(seizure.rate ~ treatment + (1|subject), df_epilepsy,REML=F)
summary(intercept_model)

slope_model <- lmerTest::lmer(seizure.rate ~ treatment + (1+base|subject), df_epilepsy,REML=F)
summary(slope_model)

# Interaction model
interaction_model <- lm(seizure.rate ~ treatment*base, data = df_epilepsy)
summary(interaction_model)

```

2.3. Do model comparison.
```{r}
# Model comparison mixed-effects models AIC 
anova(intercept_model, slope_model)
anova(logistic_model, intercept_model, slope_model, interaction_model)

# Calculating R-squared for mixed-effects model
MuMIn::r.squaredGLMM(intercept_model)
MuMIn::r.squaredGLMM(slope_model)
MuMIn::r.squaredGLMM(interaction_model)

# Calculating R-squared for logistic regression model
PseudoR2(logistic_model, c('McFadden'))

```

2.5. Write up a short summary of your model.
```
lmerTest in R is used to produce a linear mixed-effects model. The model has seizure rate as outcome variable and treatment as predictor value. As random effects, the model has random intercepts for subject to take the individual base lines into account as well as by-base random slopes. 
The outcome of the model shows that seizure rate is not significantly modulated by treatment (p > 0.05). When going from the placebo treatment to the progabide treatment, the seizure rate decreases with 0.5964 (p-value = 0.557), so even though the effect is not statistically significant it does approach significance. The whole model accounts for 74.5% of variance (R2 conditional = 0.745, and the fixed effects account for 0,075 % (R2 marginal = 0.00075). 
```

3. Exercise 2.3 in the book: Go to the folder Names and make a graph similar to Figure 2.8, but for girls. 
Hints: summarise_at(), str_sub(), pivot_longer() 
```{r}
# Load data 
allnames <- read_csv("allnames_clean.csv")

# Filter data so that it only have girls' names
girl_names_df <- allnames %>% 
  filter(sex == "F")

# Create vectors for the names of the girls 
names <- as.vector(girl_names_df$name)

# Extract last letter in names 
n_last <- 1
girl_names_df <- mutate(girl_names_df,lastletter = substr(names, nchar(names) - n_last + 1, nchar(names)))

# Move the column 'lastletter' to first position 
girl_names_df <- girl_names_df %>% 
  relocate(lastletter, .before = X1880)

# Find sum of girls' names ending in each letter
girl_names_df <- girl_names_df %>% 
  group_by(lastletter) %>% 
  summarise_at(vars(X1880:X2010), funs(sum))
girl_names_df

# Using the function pivot_longer() to make the data set longer by increasing the number of rows and decreasing the number of columns
girl_names_df <- girl_names_df %>% 
  pivot_longer(!lastletter, names_to = "year", values_to = "percentage")

# Calculate total amount of names per year 
girl_names_df <- girl_names_df %>% 
  group_by(year) %>% 
  mutate(sum_by_year = sum(percentage))

# Convert each lastletter to percentage
girl_names_df <- girl_names_df %>% 
  mutate(percentage = (percentage/sum_by_year)*100) 
# Question: How do I round percentage off? 

# Make graph 
girl_names_df %>% 
  ggplot() + 
  aes(x = year, y = percentage, group = lastletter, colour = lastletter) + 
  geom_line() +
  ggtitle("Last letters of girls' names") +
  labs(x = "Year", y = "Percentage of all girls' names that year") 
```